Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-22 ms.tex
moved result on multiple types to after result 1; changes subsections to paragraphs; formatting

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newtheorem{result}{Result}
%\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
%\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
%\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}

@@

\subsection*{Mean fitness principle}

@@

% Mean fitness principle

@@

\begin{proof}
%$\hat{\bar w}$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
%and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{\hat v_k} = 1$ and $\sum_k{\hat u_k \hat v_k} = 1$.
Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v.
\end{equation}

@@

% Reproductive value principle

An interesting intermediate result in Eq.~\ref{eq:mfp_intermediate} is

\begin{equation}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
w_k \Big(\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k\Big).
\end{aligned}
\end{equation}
Here, $\hat u_k$ can be regarded as the \emph{reproductive value} of type $k$~\citep[pg.~27]{Fisher1930}, which gives the relative contribution of type $k$ to the long-term population (see~\nameref{sec:AppB}).
Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

@@

% Threshold rule

We turn our attention to the case where the transition occurs from a subset $K$ of the types.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
If the transition rates are controlled by a threshold rule, then 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases}.
\end{equation}
If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}\label{eq:mfp_K}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)}.
\end{aligned}
\end{equation}
Thus
\begin{multline}\label{eq:mfp_sign_threshold}
sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{eq:mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

\subsection*{Evolutionary genetic stability}

@@

% Evolutionary stability

@@

Note that if the modifier has the same effect on all individuals (i.e., $C_k = \hat\alpha$ and $\tilde{C}_k = \alpha$ for all $k$) then $\cl_{ex}$ becomes
\begin{equation}
\cl_{ex} = ((1 - \alpha)I + \alpha M) D /\hat{\bar w},
\end{equation}  
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.~\ref{eq:mfp_K}, we find
\begin{equation}\label{eq:mutational_loss}
\frac{\partial \bar w}{\partial \alpha} = 
-\frac{1}{\alpha}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \bar w \Big).
\end{equation}
Eq.~\ref{eq:mutational_loss} has been derived by~\citet[eq.~24]{Hermisson2002}.
Furthermore, the term in the parentheses of Eq.~\ref{eq:mutational_loss} is the difference between the 
\emph{ancestral mean fitness} and the \emph{stable population mean fitness}, and has been termed the \emph{mutational loss} by~\citet[eq.~24]{Hermisson2002} -- "the linear response of the equilibrium mean fitness to changes in the mutation rates".
Following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$, and therefore we can conclude that the \emph{mutational loss} is positive and that the \emph{ancestral mean fitness} is higher then the \emph{stable population mean fitness}.

@@

% Conclusions
\subsection*{Conclusions}

@@

\section*{Appendices}
% Appendix A: mean fitness
%\subsection*{Appendix A}\label{sec:AppA}

%\begin{lemma}
%The normalizing factor $\bar{w}$ in Eq.~\ref{eq:model} is the population mean fitness
%\begin{equation}
%\bar{w} = \sum_k{f_k w_k}.
%\end{equation}
%\end{lemma}

%\begin{proof}
%By definition, the normalizing factor $\bar{w}$ ensures that $\sum_k{f'_k}=1$.
%Denote $e = (1, ..., 1)$ the vector with ones at all positions and rewrite $e\tr f' = 1$.
%Therefore, 
%\begin{multline}\label{eq:mean_fitness}
%\bar{w} = 
%\bar{w} e\tr f' = 
%e\tr (I - C + M C) D f = \\
%(e\tr I - e\tr C + e\tr M C) D f = \\
%(e\tr - e\tr C + e\tr C) D f = \\
%e\tr D f = 
%\diag{D} f = 
%\sum_k{f_k w_k},
%\end{multline}
%using the fact that $e\tr M = e\tr$ and $e\tr I = e\tr$ because $M$ and $I$ are stochastic matrices.
%\end{proof}

%Note that this lemma also ensures that the stable population mean fitness $\hat{\bar w}=\sum_k{v_k w_k}$ is the leading eigenvalue of $(I - C + M %C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.

>>>>

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newtheorem{result}{Result}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}

@@

% Mean fitness principle
\paragraph*{Mean fitness principle.}

@@

@@

\begin{proof}
Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v.
\end{equation}

@@

% Multiple types
\paragraph*{}
We turn our attention to the case where the transition rates from a subset $K$ of the types are correlated, that is, $C_j=C_i$ for $i,j \in K$.
In this case, the following directly results from Eq.~\ref{eq:mfp}.

\begin{corollary}
The sensitivity of the stable population mean fitness to change in the rate of transition $\tau$ from types in $K$ is
\begin{equation}\label{eq:mfp_K}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)},
\end{aligned}
\end{equation}
and
\begin{multline}\label{eq:mfp_sign_threshold}
sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big).
\end{multline}
\end{corollary}

Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated mutant alleles, $k$ (see \nameref{sec:mut_model_2}), 
and a modifier allele that increases the mutation rates in individuals with at least $\pi$ mutant alleles.
According to Eq.~\ref{eq:mfp_sign_threshold}, this modifier will increase the stable population mean fitness if individuals with $\pi$ or more mutant alleles have below-average fitness.

% Reproductive value principle
\paragraph*{Reproductive value principle.}

An interesting intermediate result in Eq.~\ref{eq:mfp_intermediate} is
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = 
w_k \Big(\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k\Big).
\end{equation}
Here, $\hat u_k$ can be regarded as the \emph{reproductive value} of type $k$~\citep[pg.~27]{Fisher1930}, which gives the relative contribution of type $k$ to the long-term population (see~\nameref{sec:AppB}).
Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

@@

% Evolutionary stability
\paragraph*{Evolutionary genetic stability.}

@@

@@

% Reduction principle & mutational loss
\paragraph*{Reduction principle and mutational loss.}

Note that if the modifier has the same effect on all individuals (i.e., $C_k = \hat\alpha$ and $\tilde{C}_k = \alpha$ for all $k$) then $\cl_{ex}$ becomes
\begin{equation}
\cl_{ex} = ((1 - \alpha)I + \alpha M) D /\hat{\bar w},
\end{equation}  
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.~\ref{eq:mfp_K}, we have
\begin{equation}\label{eq:mutational_loss}
\frac{\partial \hat{\bar{w}}}{\partial \alpha} = 
-\frac{1}{\alpha}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big),
\end{equation}
which has been derived by~\citet[eq.~24]{Hermisson2002}.
They called the term in the parentheses of Eq.~\ref{eq:mutational_loss}, which is equal to the difference between the \emph{ancestral mean fitness} $\sum_{k=1}^{n} \hat u_k \hat v_k w_k$ and the \emph{stable population mean fitness} $\hat{\bar{w}}$, the \emph{mutational loss}.
Following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$, and we can conclude that the \emph{mutational loss} is positive and that the \emph{ancestral mean fitness} is higher then the \emph{stable population mean fitness}.

@@

% Conclusions
\paragraph*{Conclusions.}

@@

\section*{Appendices}
